---
title: "Lab 1B"
name: ""
output: 
  pdf_document: 
    toc: yes
    keep_tex: yes
---

The following dataset, titled "Rain in Australia," was obtained via Kaggle.com via author Joe Young et. al.

This dataset contains daily weather observations from numerous Australian weather stations during the year 2016.  The dataset measured variables: Date, Location, MinTemp (C), MaxTemp (C), Rainfall (mm), Evaporation (mm per 24 hours), Sunshine (hours per day), WindGustDir (strongest gust in 24 hours), WindGustSpeed (km/h), WindDir9am (direction at 9 AM), ..., RainToday (Yes/No) and RainTomorrow (Yes/No).

```{r}
library(readxl)
rain =read_excel("C:\\Users\\mbles\\Documents\\weatherAUS.xls.xlsx")
rain = na.omit(rain)
head(rain, 4)
```



## Question 1
Using 1000 randomly sampled observations from the dataset `rain`, plot a scatter plot of `MaxTempF` (x) vs. `Humidity3PM` (y).  `MaxTempF` should be the `MaxTemp` variable converted to degrees F.  Color the point on the scatterplot blue if it rained that day and red if it did not rain that day.  The plot should have x- and y- axis labels in addition to a title.  After the plot, include a discussion on observations from the scatterplot.

Hints:
- To sample from `rain`, use `sample_n()` from package `dplyr`.
- To create MaxTempF, use conversion (9/5)C + 32 to change C to F.
- Create a second new column named Colors that assigns a color based on RainToday variable.

```{r}
library(dplyr)
set.seed(1000)
WeatherSample = (sample_n(weatherAUS_xls, 1000))

MaxTemp = WeatherSample[4]
MaxTempF = unlist(MaxTemp*(9/5)  + 32)#won't plot without doing unlist

Humidity3PM = unlist(WeatherSample[15])

RainedToday = unlist(WeatherSample[23])
RainedToday = ifelse(RainedToday == "No", "Blue", "Red")
#ifRainedToday[i] == "No", set it to "Blue", else set it to "Red".

plot(MaxTempF, Humidity3PM, xlab = "Highest Daily Temperature (F°)", ylab = "Humidity Level at 3pm (%)", col = RainedToday) 

title("Maximum Daily Temperature vs Afternoon Humidity")  

legend(x ="topright", legend = c("Yes", "No"), fill = c("Blue", "Red"), title = "Rain")

r = cor(as.double(MaxTempF), as.double(Humidity3PM), use = "complete.obs")  
text(40, 40, paste("r = ",round(r, 3)))
```

The scatterplot shows a weak negative correlation between the max temperatures and the humidity levels at 3pm, as can be seen by the points being somewhat clustered together forming.  If this data is a representative sample, that indicates that, in Australia, in 2016, the humidity level at 3pm was a weak predictor for the highest temperature in a given day.  Specifically, the correlation coefficient, which I added to the graph is -0.538, which, squared, gives the cooefficient of variation, which is 0.289, which means that only about 28.9% of the variation in the daily high temperature readings is predicted by the variation in the humidity readings taken at 3pm.  
Additionally, while there doesn't seem to be much of a pattern in the rainy vs non-rainy days it is worth noting that there are very few rainy days with a humidity of greater than 80% compared to non-rainy days.  


## Question 2

### Part A
Plot a histogram of `Rainfall`, including all values from `rain` dataset given that it had rained that day.  Display density on the y-axis.  Label the axes and give the graph a title.  Colorize the bars.  Include a discussion after the plot.

```{r}
    Weather = weatherAUS_xls

    RainedToday = Weather$RainToday
    
    NonzeroRainfall = ifelse(RainedToday == "Yes", Weather$Rainfall, NA)
    
    NonzeroRainfall = NonzeroRainfall[!is.na(NonzeroRainfall)]
    
    hist(NonzeroRainfall, freq = F, col = "lightblue", 
         xlab = "Rainfall(mm)", ylab = "Density", ylim = c(0,0.05),
         main = "Histogram and Density Function of Daily Rainfall \n in Austraulia in 2016", labels = T)
    
    DensityFunction = density(NonzeroRainfall, kernel = "gaussian")
    lines(DensityFunction$x, DensityFunction$y, type = 'l', lwd = 2)
    
    #Feel free to ignore the code below removing outliars.  I know it wasn't required.  I just wanted to see the plots with and without outliars and compare since the outliars were skewing things so much it was hard to see any patterns otherwise.
    
    RemoveObviousOutliars = ifelse(NonzeroRainfall < 55, NonzeroRainfall, NA)
    RemoveObviousOutliars = RemoveObviousOutliars[!is.na(RemoveObviousOutliars)]
    
    hist(RemoveObviousOutliars, freq = F, col = "lightblue", 
         xlab = "Rainfall(mm)", ylab = "Density", ylim = c(0,0.15),
         main = "Histogram and Density Function of Daily Rainfall \n in Austraulia in 2016 without obvious outliars", labels = T)
    
    DensityFunction = density(RemoveObviousOutliars, kernel = "gaussian")
    lines(DensityFunction$x, DensityFunction$y, type = 'l', lwd = 2)
    
    
    #24.9 is upper outliar boundary. Lower boundary is negative and hence irrelevant 
    RemoveAllOutliars = ifelse(NonzeroRainfall < 25, NonzeroRainfall, NA)
    RemoveAllOutliars = RemoveAllOutliars[!is.na(RemoveAllOutliars)]
    
    hist(RemoveAllOutliars, freq = F, col = "lightblue", 
         xlab = "Rainfall(mm)", ylab = "Density", ylim = c(0,0.15),
         main = "Histogram and Density Function of Daily Rainfall \n in Austraulia in 2016 without any outliars", labels = T)
    
    DensityFunction = density(RemoveAllOutliars, kernel = "gaussian")
    lines(DensityFunction$x, DensityFunction$y, type = 'l', lwd = 2)
```
There isn't really that much to say here.  The vats majority (about 89%) of the rain measurements are evidently in the range 0-50mm, and the few are incredibly spread out, in the range of around 50-225mm (note that the densities aren't actually zero; they're just so small they got rounded to zero).  It's so spread out compared to the data in the range 0-50, in fact, that it's difficult to even see most of the bars after the three leftmost ones.

### Part B
Find the mean and median of the `Rainfall` in mm for days that it rained in Australia.  Do these numbers correspond with the image in part A?

```{r}
    mean(NonzeroRainfall)
    median(NonzeroRainfall)
    
    mean(RemoveObviousOutliars)
    median(RemoveObviousOutliars)
    
    mean(RemoveAllOutliars)
    median(RemoveAllOutliars)
```
Yes, the mean and median make sense based on the distribution of the data shown in the histogram.  The data is highly right skewed with multiple extreme upper outliars.  As a result, the mean is significantly larger than the median.  Removing first the obvious outliars, then all outliars mitigates this effect greatly, though the data is still right skewed, so the mean is still larger.


## Question 3

### Part A
Plot a bar graph representing *total* `Rainfall` in 2016 by `Location`.  Make sure to label the x- and y- axes, colorize the bars, include a title, and include the appropriate response labeled above each bar.  Utilize all observations from dataset `rain`. Include a discussion after the plot.

```{r message=TRUE, warning=TRUE, paged.print=FALSE}
    library(dplyr)
    
    RainfallByLocation = tapply(Weather$Rainfall, Weather$Location, sum)
    
    RainfallByLocation = RainfallByLocation[!is.na(RainfallByLocation)]
    
    counts = as.vector(RainfallByLocation)
    
    Locations = c('Adelaide','Albury','Canberra','Coffs-Harbour','Darwin','Hobart','Launceston',
              'MB Airport','Mildura','Perth','Perth Airport','Sydney','Sydney Airport')
    
    par(mai=c(1.02,0.82,0.8,0.1)) #adjust margins so labels will fit on graph

    RainfallBarGraph = barplot(RainfallByLocation,names.arg = Locations, 
                               main = "Total 2016 Rainfall by Location", xlab = "Location", 
                               ylab = "Rainfall(mm)", ylim = c(0, 1500), xlim = c(0, 43), width = 2.5,
                               col = "deeppink", 
                               cex.names = 0.4) 
    
    text(x = RainfallBarGraph, y = counts + 90, labels = as.character(counts), cex = 0.8, col = "purple")

```
There's a decent amount of variation in the total rainfall between cities, although nothing crazy.  The main thing of note is that the graph includes only 13 total locations because the none of the other locations had rainfall data for every date and so the total rainfall for them could not be computed.


### Part B
Plot a bar graph representing *average* `Rainfall` per day based on `Location`.  Make sure to label the x- and y- axes, colorize the bars, include a title, and include the appropriate response labeled above each bar.  Utilize all observations from dataset `rain`. Include a discussion after the plot.

```{r}
    AverageRainfallByLocation = tapply(Weather$Rainfall, Weather$Location, mean)
    
    AverageRainfallByLocation = AverageRainfallByLocation[!is.na(AverageRainfallByLocation)]
    
    counts2 = as.vector(AverageRainfallByLocation)
    
    counts2
    
    Locations = c('Adelaide','Albury','Canberra','Coffs-Harbour','Darwin','Hobart','Launceston',
              'MB Airport','Mildura','Perth','Perth Airport','Sydney','Sydney Airport')
    
    par(mai=c(1.02,0.82,0.8,0.1)) #adjust margins so labels will fit on graph

    AverageRainfallBarGraph = barplot(AverageRainfallByLocation,names.arg = Locations, 
                               main = "Average 2016 Rainfall by Location", xlab = "Location", 
                               ylab = "Rainfall(mm)", ylim = c(0, 4.1), xlim = c(0, 43), width = 2.5,
                               col = "purple", 
                               cex.names = 0.4) 
    
    text(x = AverageRainfallBarGraph, y = counts2 + 0.2, labels = as.character(round(counts2, 2)), 
         cex = 0.8, col = "deeppink")
```


### Part C
Plot a bar graph representing *total* count of days rained in 2016 based on `Location`.  Make sure to label the x- and y- axes, colorize the bars, include a title, and include the appropriate response labeled above each bar.  Utilize all observations from dataset `rain`. Include a discussion after the plot.

```{r}

```



## Question 4

For locations `Brisbane`, `Melbourne`, `Perth`, and `Sydney`, plot side by side boxplots of the variable `Rainfall` for days where it rained.  The graph should include a title, x- and y-axis labels, and unique colors for the 4 boxplots.  The data should include all observations from the `rain` dataset.  Include a discussion after the plot.

```{r}
rainCity = rain[rain$Location %in% c('Brisbane', 'Melbourne', 'Perth', 'Sydney'), ]
boxplot(y~x)
```
